---
title: "DATA 624 | Predictive Analytics"
author: "Gabriella Martinez"
date: "`r Sys.Date()`"
output:
      html_document:
        toc: yes
        toc_float: yes
        theme: yeti
        highlight: kate
        font-family: "Arial"
        code_folding: hide
---

# Homework 7: Linear Regression
## Instructions

In Kuhn and Johnson do problems 6.2 and 6.3. There are only two but they consist of many parts.  Please submit a link to your Rpubs and submit the .rmd file as well.

## Packages
```{r message=FALSE, warning=FALSE, class.source = 'fold-show'}
library(AppliedPredictiveModeling)
library(caret)
library(dplyr)
library(mice)
library(corrplot)
```

### 6.2

**6.2. Developing a model to predict permeability (see Sect. 1.4) could save significant resources for a pharmaceutical company, while at the same time more rapidly identifying molecules that have a sufficient permeability to become a drug:**

**(a) Start R and use these commands to load the data:  **

```{r message=FALSE, warning=FALSE}
data(permeability)
```

**The matrix `fingerprints` contains the 1,107 binary molecular predictors for the 165 compounds, while `permeability` contains permeability response.**

```{r message=FALSE, warning=FALSE}
str(fingerprints)
```


**(b) The fingerprint predictors indicate the presence or absence of substructures of a molecule and are often sparse meaning that relatively few of the molecules contain each substructure. Filter out the predictors that have low frequencies using the `nearZeroVar` function from the `caret` package. How many predictors are left for modeling? ** 

```{r message=FALSE, warning=FALSE}
#719 low frequency instances
low_frequency <- nearZeroVar(fingerprints)

#remove low frequency columns using baser df[row,columns]
predictors <- fingerprints[,-low_frequency]

#388 predictors remaining
#dim(predictors)
```

There are 388 predictors left for modeling. 

**(c) Split the data into a training and a test set, pre-process the data, and tune a `PLS` model. How many latent variables are optimal and what is the corresponding resampled estimate of $R^2$?  **

```{r message=FALSE, warning=FALSE}
set.seed(624)

#70 30 split
split1<- sample(c(rep(0, 0.7 * nrow(permeability)), 
                  rep(1, 0.3 * nrow(permeability))))

#training split1[0] = 115 observations, test split1[1] = 49 observations 
#table(split1)

X_train <- predictors[split1 == 0,]
X_test <- predictors[split1 == 1,]

y_train <- permeability[split1 == 0]
y_test <- permeability[split1 == 1]

#PLS model 
plsTune <- train(X_train, y_train, 
                method='pls', metric='Rsquared',
                tuneLength=20, 
                trControl=trainControl(method='cv'),
                preProc=c('center', 'scale')
                )
plsTune
```

```{r message=FALSE, warning=FALSE}
plsTune$results %>% 
  dplyr::filter(ncomp == 7)
```

The best tune was found at `ncomp` = 7 with an $R^2$ value of 0.6443277.

**(d) Predict the response for the test set. What is the test set estimate of $R^2$? ** 

```{r message=FALSE, warning=FALSE}
#generate prediction using model and testing data
plsPred <- predict(plsTune, newdata=X_test)

#evaluation metrics
postResample(pred=plsPred, obs=y_test)
```

$R^2$ = 0.6443277

**(e) Try building other models discussed in this chapter. Do any have better predictive performance?**

```{r message=FALSE, warning=FALSE}
set.seed(123)

pcr_Tune <- train(X_train, y_train, 
                   method = "pcr",
                   tuneLength = 20,
                   trControl = trainControl("cv"),
                   preProc=c('center', 'scale')
)

#pcr_Tune

#generate prediction using model and testing data
pcrPred <- predict(pcr_Tune, newdata=X_test)

#evaluation metrics
postResample(pred=pcrPred, obs=y_test)
```

```{r message=FALSE, warning=FALSE}
set.seed(456)

enetGrid <- expand.grid(.lambda = c(0, 0.01, .1),
                        .fraction = seq(.05, 1, length = 20)) 

enet_Tune <- train(X_train, y_train, 
                    method = "enet",
                    tuneGrid = enetGrid,
                    trControl = trainControl("cv"),
                    preProc = c("center", "scale"))

#enet_Tune

#generate prediction using model and testing data
enetPred <- predict(enet_Tune, newdata=X_test)

#evaluation metrics
postResample(pred=enetPred, obs=y_test)
```

**(f) Would you recommend any of your models to replace the permeability laboratory experiment?**  

Each of the models run on the data perform relatively the same, of the three models run on the data, the model with the lowest RMSE was that of the Principal Component Analysis from `pcr_Tune` with an RMSE of 13.9929634.

### 6.3

**6.3. A chemical manufacturing process for a pharmaceutical product was discussed in Sect. 1.4. In this problem, the objective is to understand the relationship between biological measurements of the raw materials (predictors), measurements of the manufacturing process (predictors), and the response of product yield. Biological predictors cannot be changed but can be used to assess the quality of the raw material before processing. On the other hand, manufacturing process predictors can be changed in the manufacturing process. Improving product yield by 1 % will boost revenue by approximately one hundred thousand dollars per batch:**  

**(a) Start R and use these commands to load the data:**

```{r message=FALSE, warning=FALSE}
data(ChemicalManufacturingProcess)
```

**The matrix `processPredictors` contains the 57 predictors (12 describing the input biological material and 45 describing the process predictors) for the 176 manufacturing runs. `yield` contains the percent yield for each run.  **


**(b) A small percentage of cells in the predictor set contain missing values. Use an imputation function to fill in these missing values (e.g., see Sect. 3.8).**


```{r message=FALSE, warning=FALSE}
#column ids with the NAs
#unique(which(is.na(ChemicalManufacturingProcess), arr.ind = TRUE)[,2])

imputed_data <- mice(ChemicalManufacturingProcess, printFlag=F, method="cart", seed = 1)

full_data <- complete(imputed_data)

#sum(is.na(full_data))
```


**(c) Split the data into a training and a test set, pre-process the data, and tune a model of your choice from this chapter. What is the optimal value of the performance metric?  **


```{r message=FALSE, warning=FALSE}
#8 low frequency instances
low_values <- nearZeroVar(full_data)

#remove low frequency columns using baser df[row,columns]
chem_predictors <- full_data[,-low_values]

#70 30 split
split2 <- sample(c(rep(0, 0.7 * nrow(chem_predictors)), 
                  rep(1, 0.3 * nrow(chem_predictors))))

#split the data
chem_train <- chem_predictors[split2 == 0,]
chem_test <- chem_predictors[split2 == 1,]

#PLS model 
chem_pls <- train(Yield~., chem_train, 
                method='pls', metric='Rsquared',
                tuneLength=30, 
                trControl=trainControl(method='cv'),
                preProc=c('center', 'scale')
                )

chem_pls

#best fit metrics
chem_pls$results %>% 
  filter(ncomp==4)
```



**(d) Predict the response for the test set. What is the value of the performance metric and how does this compare with the resampled performance metric on the training set?  **


```{r message=FALSE, warning=FALSE}
#predict
chem_predict = predict(chem_pls,chem_test)

postResample(pred = chem_predict, obs =chem_test$Yield)

```

The $R^2$ value raised slightly on the `chem_test$Yield` data suggesting the model performed better using test set than on the training set. 


**(e) Which predictors are most important in the model you have trained? Do either the biological or process predictors dominate the list?**

Using the `varImp()` function, it is found that the ManufacturingProcess predictors are slightly more important as they make up 12 of the top 20 most important predictors followed by 8 BiologicalMaterial predictors.

```{r message=FALSE, warning=FALSE}
varImp(chem_pls)
```


**(f) Explore the relationships between each of the top predictors and the response. How could this information be helpful in improving yield in future runs of the manufacturing process?  **

Using the correlation matrix below, the `Yield` response variable doesn't have strong correlations with many of the important predictors identified in the previous step. The strongest positive correlations are with `ManufacturingProcess32` and `ManufacturingProcess09` with correlation values 0.61 and 0.50 respectively. The strongest negative correlations are with `ManufacturingProcess36` and `ManufacturingProcess13` with correlations values -0.52 and -0.50, respectively. The response variable does not have strong correlations with the important Biological Material predictors from part e. 

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
corr_vals <- chem_predictors %>% 
  select('Yield', 'ManufacturingProcess32','ManufacturingProcess36',
         'ManufacturingProcess13','ManufacturingProcess17',
         'ManufacturingProcess09','ManufacturingProcess12',
         'ManufacturingProcess11','ManufacturingProcess33',
         'BiologicalMaterial02', 'BiologicalMaterial08',
         'BiologicalMaterial06', 'BiologicalMaterial12')

corr_plot_vals <- cor(corr_vals)

corrplot.mixed(corr_plot_vals, tl.col = 'black', tl.pos = 'lt', 
         upper = "number", lower="circle")
```


<!------- Below is for removing excessive space in Rmarkdown | HTML formatting -------->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>